### 2017-07-05 paper/app.tex
re-shuffled the proof of the main theorem a bit

\section{Details of Program Syntax}
In this subsection we present the details of the syntax of (affine) 
probabilistic 
programs.

@@

>>>>

%\nouppercaseheads
%\def\@secfont{\sffamily\bfseries}
\section{Details of Program Syntax}
In this subsection we present the details of the syntax of (affine) 
probabilistic 
programs.

@@

\def\xx{\ref{prop:conditional-exp-existence}}

\section{Proof of Proposition~\xx}

We first recall the general statement of the Radon-Nikodym theorem. Given two measurable spaces\footnote{A generalization of a probability space where the measure of $\Omega$ does not have to be 1, but any non-negative number or even infinity.} $(\Omega,\genfilt,\mu)$ and $(\Omega,\genfilt,\nu)$, we say that $\nu$ dominates $\mu$, written $\mu<<\nu$ if for all $A\in \genfilt$, $\nu(A)=0$ implies $\mu(A)=0$. Radon-Nikodym theorem states that if both $\mu$ and $\nu$ are sigma-finite (that is, $\Omega$ is a union of countably many sets of finite measure under $\nu$ and $\mu$), then $\mu<<\nu$ implies that there exists an almost-surely unique $\genfilt$-measurable function $f\colon \Omega\rightarrow [0,\infty)$ such that for each $A \in\genfilt$, the Lebesgue integral of the function $f\cdot\indicator{A}$ in measurable space  $(\Omega,\genfilt,\nu)$ is equal to $\mu(A)$. The function $f$ is called a Radon-Nikodym derivative of $\mu$ w.r.t. $\nu$, and we denote in by $\frac{d\mu}{d\nu}$.

Now assume that $X$ is a non-negative real-valued random variable in some probability space $(\Omega,\genfilt,\probm)$ and $\genfilt'$ is a sub-sigma algebra of $\genfilt$. Note that $\probm$ is sigma-finite. Define a measure $\mu$ on $\genfilt'$ by putting $\mu(A)=\E[X\cdot\indicator{A}]$, for each $A\in \genfilt$ (here $\E$ is the expectation operator, i.e. the Lebesgue integral, in probability space $(\Omega,\genfilt',\probm')$, where $\probm'$ is a restriction of $\probm$ to $\genfilt'$). Then $\mu$ is sigma-finite: indeed, for any $n\in \Nset$ let $A_n = \{\omega\in\Omega\mid X(\omega)\leq n\}$. Then $\mu(A_n)\in [0,n]$, in particular it is finite, and since $X$ is real-valued, we have $\Omega=\bigcup_{n=1}^{\infty} A_n$. Hence, $\frac{d\mu}{d\probm'}$ exists and is almost-surely unique. It is now easy to check that $\frac{d\mu}{d\probm'}$ satisfies the condition defining the conditional expectation $\E[X\mid \genfilt']$: indeed, the condition is equivalent to $\E[X\mid \genfilt']$ being a derivative of $\mu$ w.r.t. $\probm'$. This concludes the proof.

%%\section{Integrability of Variables in Non-Affine Programs}
%%
%%Consider the o

\section{Computations for the proof of Theorem~\ref{THM:LEXRSM-MAIN}}

Recall that we aim to prove equation~\eqref{eq:lexrsm-soundness-main}.

For 
$k=\fixn{i}$ the sum on the right-hand side equals $0$, so the 
inequality immediately follows from the definition of $Y_k$. Now assume 
that~\eqref{eq:lexrsm-soundness-main} holds for some $k\geq \fixn{i}$. We have that 
\begin{align}
\label{eq:lexrsm-ind-1}
\E[Y_{k+1}] &= \underbrace{\E[Y_{k+1}\cdot \indicator{\Omega\setminus D}]}_{=0=\E[Y_k\cdot\indicator{\Omega\setminus D}]} +  \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}]}_{=\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}]} + \underbrace{\E[Y_{k+1}\cdot \indicator{D \cap \{U > k\}}]}_{=\E[\vecseq{X}{k+1}{\fixn{j}}\cdot\indicator{D\cap \{U>k \} }]},
\end{align}
where the equality $\E[Y_{k+1}\cdot \indicator{D \cap \{U \leq k\}}] =\E[Y_k\cdot\indicator{D\cap \{U\leq k\}}] $ follows from the fact that $Y_{k+1}(\omega)=Y_k({\omega})=\vecseq{X}{U(\omega)}{\fixn{j}}(\omega)$ for $\omega\in \{U\leq k\}$, and similarly for the last term. We prove that 
\begin{equation}
\label{eq:lexrsm-ind-2}
\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \}} -\eps\cdot \indicator{D\cap \{U>k\} \cap \{\levelrank{}{k}= \fixn{j}\} }] .
\end{equation}
Indeed, it holds
\begin{equation}
\label{eq:lexrsm-ind-3}
\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \}}] = \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }]  + \E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j}\} }], 
\end{equation}
since $\levelrank{\omega}{k}\geq \fixn{j}$ for all $\omega \in \{U>k\}$. Since the set $D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}$ is $\genfilt_k$-measurable, we get
\begin{align}
\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] &= \E[\E[\vecseq{X}{k+1}{\fixn{j}}\mid \genfilt_k]\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \} }] \label{eq:lexrsm-ind-4}\\
&\leq \E[(\vecseq{X}{k}{\fixn{j}} - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-5}\\
&=\E[(Y_k - \eps)\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}=\fixn{j} \}}] \label{eq:lexrsm-ind-6},
\end{align}
where~\eqref{eq:lexrsm-ind-4} follows from the definition of conditional expectation~\eqref{eq:cond-exp},~\eqref{eq:lexrsm-ind-5} follows from the definition of $\{\levelrank{}{k}=\fixn{j}\}$, and~\eqref{eq:lexrsm-ind-6} holds since $Y_k(\omega)=\vecseq{X}{k}{\fixn{j}}(\omega)$ for $\omega$ with $U(\omega)> k$. Almost identical argument shows that
\begin{equation}
\label{eq:lexrsm-ind-7}
\E[\vecseq{X}{k+1}{\fixn{j}}\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \} }] \leq \E[Y_k\cdot \indicator{D\cap \{U > k \} \cap \{\levelrank{}{k}>\fixn{j} \}}].
\end{equation}
Plugging~\eqref{eq:lexrsm-ind-6} and~\eqref{eq:lexrsm-ind-7} into~\eqref{eq:lexrsm-ind-3} yields~\eqref{eq:lexrsm-ind-2}. Now we can plug~\eqref{eq:lexrsm-ind-2} into~\eqref{eq:lexrsm-ind-1} to get
\begin{align}
\E[Y_{k+1}]&\leq \E[Y_k] - \eps\cdot \E[\indicator{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} }] = \E[Y_k] - \eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \} ) \nonumber\\
&\leq \fixn{B}\cdot \probm(D) - \eps\cdot\left(\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D 
\cap \{U\geq k\} \cap \{\noofdec_k = \ell\})\right) -\eps\cdot \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}),
\end{align}
where the last inequality follows from induction hypothesis. Hence, using $D_{k,\ell}$ as a shorthand for $D 
\cap \{U\geq k\} \cap \{\noofdec_k = \ell\}$, to prove~\eqref{eq:lexrsm-soundness-main} it remains to show that
\begin{equation}
\label{eq:lexrsm-ind-8}
\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}) + \probm(D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j} \}) = \sum_{\ell=0}^{k+1-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}).
\end{equation}
The left-hand side of~\eqref{eq:lexrsm-ind-8} is equal to
\begin{align}
&\phantom{+}\;\sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell} \cap \{\levelrank{}{k}=\fixn{j}\}) + \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )\nonumber \\ 
&+\sum_{\ell=0}^{k-\fixn{i}}\probm(\underbrace{D \cap \{U>k\} \cap \{\levelrank{}{k} = \fixn{j}\} \cap \{\noofdec_k = \ell\}}_{=D_{k,\ell} \cap \{\levelrank{}{k} = \fixn{j}\}})
\nonumber \\
&= \sum_{\ell=0}^{k-\fixn{i}}(\ell+1) \cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k} = \fixn{j}\}) +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \nonumber\\
&=\sum_{\ell=0}^{k-\fixn{i}} (\ell+1)\cdot \probm{(D_{k+1,\ell+1} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} ) \label{eq:lexrsm-ind-9}\\
&= \sum_{\ell=1}^{k+1-\fixn{i}} \ell\cdot \probm{(D_{k+1,\ell} \cap \{\levelrank{}{k} = \fixn{j}\})} +  \sum_{\ell=0}^{k-\fixn{i}} \ell\cdot\probm(D_{k+1,\ell}\cap \{\levelrank{}{k}>\fixn{j}\} )
\nonumber \\
&= (k+1-\fixn{i})\cdot \probm{(D_{k+1,k+1-\fixn{i}}\cap\{\levelrank{}{k} =\fixn{j} \} )} + \sum_{\ell=1}^{k-\fixn{i}} \ell \cdot \probm(D_{k+1,\ell})  \nonumber\\
&= (k+1-\fixn{i})\cdot\probm(D_{k+1,k+1-\fixn{i}}) + \sum_{\ell=1}^{k-\fixn{i}}\ell\cdot\probm(D_{k+1,\ell}) = \text{ right-hand side of~\eqref{eq:lexrsm-ind-8}} \label{eq:lexrsm-ind-11}.
\end{align}
The individual steps in the above computation are justified as follows: in~\eqref{eq:lexrsm-ind-9} we use the facts that for all $\omega$'s whose level in step $k$ is $\fixn{j}$ it holds that $\noofdec_k(\omega)+1=\noofdec_{k+1}(\omega)$, and similarly, for $\omega$'s whose level in step $k$ is $>\fixn{j}$ it holds $\noofdec_k(\omega)=\noofdec_{k+1}(\omega)$. Moreover, for all $\omega\in D_{k,\ell}$ it holds that $\levelrank{k}{\omega}\geq \fixn{j} \Rightarrow U(\omega)\geq k+1$. Finally, in~\eqref{eq:lexrsm-ind-11} we use the fact that all $\omega\in D_{k+1,k+1-\fixn{i}}$ need to have level $\fixn{j}$ in step $k$, since otherwise such an $\omega$ would need to have level $\fixn{j}$ for at least $k+1-\fixn{i}$ times within steps $\{\fixn{i},\fixn{i}+1,\dots,k-1\}$, but there are $k-\fixn{i}$ such steps, a contradiction. This concludes the proof of~\eqref{eq:lexrsm-soundness-main}.
